---
title: "CS 448B Assignment 2"
author: "Joanne Jang"
date: "April 17, 2016"
output: html_document
---

## Why I chose this dataset
Among all the presents (and love) I received on my 21st birthday earlier this
month, my favorite was the Inside Airbnb dataset Christine (who's also taking
this class) gave me. I love Airbnb-- its mission and the company culture-- but
I'm not well-informed re: the controversy surrounding their practices, so I
thought that exploring this dataset would be a great way to learn more about
the controversy while sifting through (daydreaming about?) listings in NYC.

## About the dataset
The subset of the huge dataset I chose consists of review metadata, listings,
and neighborhood data in New York City, NY (35957 listings) compiled on 2/2/2016.
As I hinted earlier, the website hosting the dataset-- insideairbnb.com-- is
biased; it dubs Airbnb as "illegal hotels," and I was first hesitant to use this
for my project fearing that the compilation of the data itself might have been
biased. Upon reading the disclaimers and articles written about it, however, I
decided to proceed because the compilation itself seemed harmless and it
sounded too exciting to resist.

## Initial question
Are most Airbnb listings in NYC instances of "home-sharing" (spare-room rentals)
or "illegal hotels" as its critics claim? In other words, what percentage of
listings are homes that are rented out in its entirety?

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(leaflet)

listings <- read_csv("~/Desktop/taking/CS448B/Airbnb/listings.csv")
reviews <- read_csv("~/Desktop/taking/CS448B/Airbnb/reviews.csv")
neighborhoods <- read_csv("~/Desktop/taking/CS448B/Airbnb/neighbourhoods.csv")
```

## Pre-exploration
```{r}
summary(listings)
listings %>% head()
reviews %>% head()
neighborhoods %>% head()
```

I see that neighborhood information has already been joined to provide
neighborhood group information in listings, so I don't need to consider
neighborhoods anymore. I also don't think I will be using reviews metadata (at
least for now), so that leaves me with one (rich) data frame to work with:
listings.

Some columns that will come in handy in answering my initial question are 
`room_type` and `availability_365`. I realized that there's no actual booking
information (e.g. how popular a certain listing is), which isn't completely 
necessary in answering my question, but would be good to have as additional
data. The readme page suggests a review rate of 50% to convert reviews to
estimated bookings (e.g. if there are 2 reviews, we can estimate the # of
actual bookings to be 4), and I'll use that.

```{r}
listings <- listings %>% mutate(estimated_bookings = 2*number_of_reviews)
```

#### 1. Around what percentage of all listings are entire home/apts?

```{r}
listings %>% group_by(room_type) %>%
    summarise(percentage = n()/nrow(listings), n =n())
```

Around 53.6% of all listings are for entire home/apts, followed by 43.3% in
private rooms and 3.1% in shared rooms.

To visualize this with a bar chart:
```{r}
listings %>% group_by(room_type) %>%
    summarise(n = n()) %>%
    ggplot(mapping = aes(x = room_type, y = n, fill = room_type)) +
    geom_bar(stat = "identity") +
    labs(x = "", y = "Number of Listings", title = "Room Types of Airbnb Listings in NYC") +
    theme(legend.position = "none")
```

I should take into account the availability of the listings as well; if, say,
an entire home is listed as available for 365 days, we know that the host does
not live in that home, and more of those instances would lead Airbnb further
away from the spare-room-rentals-for-homeowners-looking-to-make-a-few-extra-
bucks value proposition.

#### 2. Are there hosts actually running businesses off of Airbnb?

I want to examine the distribution of the # of listings each host has on
Airbnb. I saw that there was a column called `calculated_host_listings_count`,
but want to verify by computing it myself using the other given columns.

```{r}
listings %>% group_by(host_id) %>%
    summarize(my_host_listings_count = n()) %>%
    arrange(desc(my_host_listings_count)) %>%
    head()

listings %>% select(host_id, calculated_host_listings_count) %>%
    distinct() %>%
    arrange(desc(calculated_host_listings_count)) %>%
    head()
```

Whoa, the numbers are pretty different. My guess is that I'm not considering
some factors `calculated_host_listings_count` considered when it was compiled?

Let's zoom in on host id 30283594 that appeared as 1st in both versions but had
a different number of listings (56 vs. 41):

```{r, results = "hide"}
listings %>% filter(host_id == 30283594)
```

Piping the above command to View(), I see that some `host_name` entries are
n/a even though they have the same `host_id`. Could host_name == NA matter?

```{r, results = "hide"}
listings %>% filter(host_id == 30283594, !is.na(host_name)) %>% nrow()
```

That was it! I'm guessing the person who compiled this data set took out all the
missing values before calculating the host listings count, so listings for which
`host_name` was n/a didn't get counted. I think they should be counted, since
the other columns seem to indicate that the entries are valid, so I will be
using `my_host_listings_count`.

```{r}
listings <- listings %>% group_by(host_id) %>%
    mutate(my_host_listings_count = n())
```

```{r}
listings %>% group_by(my_host_listings_count) %>%
    mutate(listings_count_tier = ifelse(my_host_listings_count == 1, "1",
                                    ifelse(my_host_listings_count == 2, "2",
                                    ifelse(my_host_listings_count == 3, "3",
                                           "4 or more")))) %>%
    ggplot(mapping = aes(x = listings_count_tier)) +
    geom_bar(stat = "count") +
    labs(x = "# of Listings Per Host", y = "# of Hosts",
         title = "Listings Count Per Host")
```

To answer the question, yes, there seems to be a non-trivial number of hosts
with three or more listings on Airbnb (with the max number of listings being 56
as shown above), which goes against the spare-room rental value proposition I
mentioned earlier. Another way to check whether or not a number of Airbnb
listings go against the value proposition is to look at the # of days available
for each listing; if the listing's `room_type` is an entire home and it's
listed as available for 365 days, we definitely know that the host doesn't live
in the home.

#### 3. What is the distribution of availability of the listings in a given 365-
day period?

I'd expect for it to be bimodal, with the maxima near both ends of the
availability spectrum.

```{r}
listings %>% ggplot(mapping = aes(x = availability_365)) +
    geom_histogram(bins = 30) +
    labs(x = "", y = "Number of Listings",
    title = "Histogram of Availability of Airbnb Listings in a Given 365-day Period") +
    scale_x_continuous(breaks = seq(0, 365, 30)) 
```

Turns out it's true. I'd like to look at the two spikes within the histogram:

```{r}
listings %>% filter(availability_365 < 10 | availability_365 > 355) %>%
    group_by(availability_365) %>%
    summarise(n = n())
```

There are 6209 listings that are available 365 days a year (followed by 611
listings for 363 days and 700 listings for 364 days). Now I'm curious what the
1 or 2 off-days are for the houses listed as available for 363 and 364 days;
could day be a specific holiday (e.g. New Year's) when the host would like to
stay at their place for once?

But perhaps more importantly:

#### 4. Why are there 6947 listings for `availability_365` = 0? What does it
mean for rooms to be available for 0 days?

```{r, results = 'hide'}
listings %>% filter(availability_365 == 0) %>% head()
```

At first glance, just looking at the head, I notice that the rows don't seem to
have anything particular in common other than neighbourhood, but this is
probably because it's implicitly arranged by neighbourhood. Could
`availability_365` have to do with neighbourhood/neighbourhood_group?

```{r}
listings %>% mutate(`0 Days Available` = (availability_365 == 0)) %>%
    group_by(neighbourhood_group) %>%
    ggplot(mapping = aes(x = neighbourhood_group, fill = `0 Days Available`)) +
    geom_bar(stat = "count") +
    labs(x = "Borough", y = "Number of Listings per Borough", title = "Zero vs. Non-zero Days of Availability for Listings")
```

Around 20% of listings in Brooklyn and Manhattan have availability 0, as do 13%
of listings in Queens and 6-7% of listings in Bronx and Staten Island. Seeing
the total numbers, though, I realize I should differentiate between popular
tourist destinations and mostly residential regions. Anyway this doesn't tell
us much about the 0 available days.

Going back to the documentation, I found this tidbit: "The Airbnb calendar for a
listing does not differentiate between a booked night vs an unavailable night,
therefore these bookings have been counted as "unavailable". This serves to
understate the Availability metric because popular listings will be "booked"
rather than being "blacked out" by a host."

This is significant. Theoretically, a house could be available for 365 days but
if it's totally booked for 363 of those days, `availability_365` would be 2.
This would explain the 6947 listings figure (for which the availability variable
is 0); if the house is totally booked, regardless of how many days it was
initially listed as available, `availability_365` is 0. I wrongly assumed that
'not available' meant blocked out by the host.

#### 5. Are there more listings in popular tourist destinations?

I think there would be the most number of listings in residential areas near
popular tourist destinations.

```{r}
leaflet(listings) %>%
  addTiles() %>%
  addMarkers(clusterOptions = markerClusterOptions())
```

Zooming in on the cluster with the biggest count at each level, we are
ultimately taken to the blocks around Tompkins Square Park, which seems to be
a primarily residential area. I wish I knew more about New York City; all I
know about the boroughs is that the majority of the most well-known attractions
are located in Manhattan. Maybe I can use `minimum_nights` to get a rough idea
of which boroughs tend to be more residential (since the bigger the
`minimum_nights` value, the more likely that the host is using Airbnb to rent
out his/her place)?

#### 6. What is the distribution of minimum nights for each borough?

```{r}
listings$minimum_nights %>% summary()
```

I'd expect for the distribution to be heavily skewed to near 0. I'd also expect
for the median values for minimum nights to be higher in less touristy (/more
residential) areas such as Queens since more people would be using it for 
longer-term rentals. Too many outliers (especially an insane max value like
1250) would flatten out the boxplot, so I'm going to filter out listings for
which `minimum_nights` > 15 to zoom in on the boxes themselves.

```{r}
listings %>% filter(minimum_nights < 15) %>%
    ggplot(mapping = aes(x = factor(neighbourhood_group),
                                  y = minimum_nights)) +
    geom_boxplot() +
    scale_y_continuous(breaks = seq(1, 20, 1)) +
    labs(x = "Borough", y = "Minumum Nights",
        title = "Distribution of Minimum Nights Per Borough")
```

It turns out I was wrong; Bronx, Queens, and Staten Island actually have the
lowest median values (1) for `minimum_nights`, as opposed to Brooklyn and
Manattan's median of 2 nights. As spineless as I may sound, I can actually
think of a number of explanations for this, what I think is the strongest
explanation being that people wouldn't resort to Airbnb to find longer-term
housing in Queens where the platform takes a portion of the rent each night;
people don't mind paying the extra fee when quality short-term housing
(especially as alternative to hotels) is extremely hard to find, as is the case
for Brooklyn and Manhattan. Which brings me to price.

#### 7. What is the distribution of price for each borough?

I filtered out the prices > $750 a night to make the boxes more prominent, but
the number of outliers especially for Brooklyn and Manhattan is pretty telling
about prices in those two boroughs.

```{r}
listings %>% filter(price < 750) %>%
    group_by(neighbourhood_group) %>%
    ggplot(mapping = aes(x = room_type, y = price, fill = room_type)) +
    geom_boxplot() +
    facet_grid(~ neighbourhood_group) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(breaks = seq(0, 750, 50)) +
    labs(x = "Room Type", y = "Price Per Night ($)",
        title = "Price Per Night of Airbnb Listings in NYC") +
    theme(legend.position = "none")
```

#### 8. Roughly how much have hosts made by using Airbnb?

```{r}
listings %>% group_by(host_id) %>%
    mutate(estimated_earning = estimated_bookings*price*minimum_nights) %>%
    filter(estimated_earning < 10000) %>%
    ggplot(mapping = aes(x = factor(neighbourhood_group), y = estimated_earning)) +
    geom_boxplot() +
    scale_y_continuous(minor_breaks = NULL, breaks = seq(0, 10000, 1000))
```

