---
title: "EDA on Sleep, Legal Drugs, and Exam Stress"
author: "Joanne Jang"
date: "5/23/2017"
output: html_document
---
data from: http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0109490#pone.0109490-Zunhammer1

```{r, message = FALSE}
# Libraries
library(readr)
library(dplyr)
library(ggplot2)

# Dataset
df <- read_delim("/Users/jjang/sleep_eda/Exam_Stress_Zunhammer_et_al_2013.csv", delim=";", col_types = cols(PSQI_ITEM1_Bedtime = col_character(), PSQI_ITEM3_Waketime = col_character(), PSQI_Time_in_Bed = col_number()))

# filtering dropouts
df <- df %>% filter(filter_dropouts == 1)
summary(df$time)
df$time <- ordered(df$time, level = c(1, 2, 3), labels=c("pre-exam", "during", "post-exam"))
df$PSQI_SUB_Sleep_Efficiency <- ordered(df$PSQI_SUB_Sleep_Efficiency, levels=c(0, 1, 2, 3), labels=c("> 85%", "75-84%", "65-74%", "< 65%"))
df$sex <- factor(df$sex, levels=c(0,1), labels=c("female", "male"))

```
#### Perceived stress
```{r}

df %>% ggplot(aes(x=factor(time), y=PSQ_total)) +
    geom_boxplot() +
    ggtitle("Perceived Stress") +
    labs(x = "exam season", y = "perceived stress score (PSQ-20)")
```

#### Perceived sleep quality index
```{r}

df %>% ggplot(aes(x=factor(time), y=PSQI_total)) +
    geom_boxplot() +
    geom_hline(yintercept=5, color="red") +
    ggtitle("Perceived Sleep Quality Index") +
    labs(x = "exam season", y = "PSQI score")
```


#### Sleep time and efficiency comparison
```{r}
df %>% ggplot(aes(x=factor(time), y=PSQI_ITEM4_Sleep_Time)) +
    geom_boxplot() +
    ggtitle("Sleep Time Distribution") +
    labs(x = "exam season", y = "sleep time (h)") +
    scale_y_continuous(breaks=seq(2,12,0.5))

summary(df$course_of_study)
```

Sleep efficiency:

# hours slept / # hours spent in bed * 100

>85%: 0
75-84%: 1
65-74%: 2
< 65: 3

The lower the score, the more efficient.

```{r}
df %>% subset(!is.na(PSQI_SUB_Sleep_Efficiency)) %>%
    ggplot(aes(x=PSQI_SUB_Sleep_Efficiency)) +
    geom_bar(stat="count", aes(fill=factor(PSQI_SUB_Sleep_Efficiency))) +
    ggtitle("Sleep Efficiency Distribution") +
    facet_grid(. ~ time) +
    labs(x = "sleep efficiency", y = "count") +
    theme(legend.position="none")
```
```{r}
df %>%
    ggplot(aes(x=PSQ_total, y=PSQI_total, color = time)) +
    geom_point() +
    geom_smooth(method='lm',formula=y~x) +
    geom_hline(yintercept=5, color="red") +
    ggtitle("Perceived Stress and Sleep Quality") +
    labs(x = "perceived stress (PSQ-20)", y = "PSQI score") +
    facet_grid(~time)
```

```{r}
df %>%
    ggplot(aes(x=PSQ_total, y=Sleep_Efficiency, color = time)) +
    geom_point() +
    geom_smooth(method='lm',formula=y~x) +
    ggtitle("Perceived Stress and Sleep Efficiency") +
    labs(x = "perceived stress (PSQ-20)", y = "sleep efficiency") +
    facet_grid(~time)
```
#### Legal drugs and sleep latency
```{r}
df %>% ggplot(aes(x= caffeine_per_day, y=PSQI_ITEM2_Sleep_Latency)) +
    geom_point() +
    geom_smooth() +
    scale_y_continuous(limit = c(0,100), breaks=seq(0, 100, 10)) +
    scale_x_continuous(limit = c(0, 5), breaks = seq(0, 5, 0.5))
    
```

There doesn't seem to be any correlation between caffeine consumption and sleep latency.


```{r}
df %>% subset(!is.na(caffeine_per_day)) %>%
    ggplot(aes(x=caffeine_per_day, y=PSQI_total, color = time)) +
    geom_point() +
    geom_smooth(method='lm',formula=y~x) +
    geom_hline(yintercept=5, color="red") +
    ggtitle("Caffeine Consumption") +
    labs(x = "caffeine per day (caffeine units)", y = "PSQI score") +
    facet_grid(~time)
```





```{r}
df %>% subset(!is.na(nicotione_per_day)) %>%
    ggplot(aes(x=nicotione_per_day, y=PSQI_total, color = sex)) +
    geom_point() +
    geom_smooth(method='lm',formula=y~x) +
    geom_hline(yintercept=5, color="red") +
    ggtitle("Nicotine Consumption") +
    labs(x = "nicotine per day (cigarettes)", y = "PSQI score") +
    facet_grid(~time)

```

 "Increased cigarette consumption was associated with decreased sleep quality."

```{r}
df %>% subset(!is.na(alcohol_drinks_per_week)) %>%
    ggplot(aes(x=alcohol_drinks_per_week, y=PSQI_total, color = sex)) +
    geom_point() +
    geom_smooth(method='lm',formula=y~x) +
    geom_hline(yintercept=5, color="red") +
    ggtitle("Alcohol Consumption") +
    labs(x = "alcohol drinks per week (standard drinks)", y = "PSQI score") +
    facet_grid(~time)
```